---
title: "final_project"
author: "Diallo Mamadou Korka"
date: "2023-12-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the libraries required for this project
```{r }
library(ggplot2)
library(tidyverse)
library(readr)
library(corrplot)
library(cowplot)
library(viridis)
library(tibble)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```

# Load the dataset
```{r}
insurance_data <- read_csv("insurance.csv")
head(insurance_data)
```
## Data Exploration
```{r} 
data_shape <- dim(insurance_data)

# Print the shape of the dataset
cat("Number of rows:", data_shape[1], "\n")
cat("Number of columns:", data_shape[2], "\n") 
```
#Data Cleaning
## cheack missing values
```{r}
colSums(is.na(insurance_data))

```
## check duplicate rows
```{r}

duplicate_rows <- sum(duplicated(insurance_data))
print(duplicate_rows)
```
## drop Duplicated rows
```{r}
data <- distinct(insurance_data)
```
#Exploring the dataset
```{r} 

# Age
age_mean <- mean(insurance_data$age)
age_sd <- sd(insurance_data$age)
age_min <- min(insurance_data$age)
age_max <- max(insurance_data$age)

# Sex
sex_counts <- table(insurance_data$sex)

# BMI
bmi_mean <- mean(insurance_data$bmi)
bmi_sd <- sd(insurance_data$bmi)
bmi_min <- min(insurance_data$bmi)
bmi_max <- max(insurance_data$bmi)

# Children
children_mean <- mean(insurance_data$children)
children_sd <- sd(insurance_data$children)
children_min <- min(insurance_data$children)
children_max <- max(insurance_data$children)

# Smoker
smoker_counts <- table(insurance_data$smoker)

# Region
region_counts <- table(insurance_data$region)

# Charges
charges_mean <- mean(insurance_data$charges)
charges_sd <- sd(insurance_data$charges)
charges_min <- min(insurance_data$charges)
charges_max <- max(insurance_data$charges)

# Print the results
cat("Age:\n")
cat("Mean:", age_mean, "\n")
cat("Standard Deviation:", age_sd, "\n")
cat("Min:", age_min, ", Max:", age_max, "\n\n")

cat("Sex:\n")
cat("Male:", sex_counts["male"], "instances\n")
cat("Female:", sex_counts["female"], "instances\n\n")

cat("BMI (Body Mass Index):\n")
cat("Mean:", bmi_mean, "\n")
cat("Standard Deviation:", bmi_sd, "\n")
cat("Min:", bmi_min, ", Max:", bmi_max, "\n\n")

cat("Children:\n")
cat("Mean:", children_mean, "\n")
cat("Standard Deviation:", children_sd, "\n")
cat("Min:", children_min, ", Max:", children_max, "\n\n")

cat("Smoker:\n")
cat("Yes:", smoker_counts["yes"], "instances\n")
cat("No:", smoker_counts["no"], "instances\n\n")

cat("Region:\n")
cat("Southwest:", region_counts["southwest"], "instances\n")
cat("Southeast:", region_counts["southeast"], "instances\n")
cat("Northwest:", region_counts["northwest"], "instances\n")
cat("Northeast:", region_counts["northeast"], "instances\n\n")

cat("Charges:\n")
cat("Mean:", charges_mean, "\n")
cat("Standard Deviation:", charges_sd, "\n")
cat("Min:", charges_min, ", Max:", charges_max, "\n")

```
#Exploring data Analysis

## information about the dataset
```{r} 
str(data)
```
#### distribution on age
```{r} 
# Create a histogram
ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution") +
  theme_minimal()
```
```{r}
# Define a function to categorize age
age_category <- function(age) {
  age_dict <- c('0-9', '10-19', '20-29', '30-39', '40-49', '50-59', '60-69', '70-79', '80-89', '90-99', '100-200')
  return(age_dict[findInterval(age, seq(0, 110, by = 10))])
}
# Print counts by age category
print("Age by category")
by_age <- table(insurance_data$cage)
print(by_age)
cat(rep('-', 80), '\n')
```
#BMI Distribution
```{r} 
ggplot(data, aes(x = bmi)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 18.5, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 24.9, color = "red", linetype = "dashed") +
  labs(title = "BMI Distribution") +
  theme_minimal()
```
```{r}
# Define a function to categorize BMI
bmi_category <- function(bmi) {
  if (bmi < 18.5) {
    return('under-weight')
  } else if (bmi >= 18.5 && bmi <= 24.9) {
    return('normal-weight')
  } else if (bmi >= 25 && bmi <= 29.9) {
    return('over-weight')
  } else if (bmi >= 30.0) {
    return('obese')
  } else {
    return('unknown')  # Handle cases where BMI is not in any category
  }
}

# Print counts by BMI category
print("BMI")
by_bmi <- table(insurance_data$cbmi)
print(by_bmi)
cat(rep('-', 80), '\n')
```

```{r}
# Create a count plot with labels
ggplot(data, aes(x = factor(children))) +
  geom_bar(fill = "skyblue", color = "black", alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Count of Children") +
  theme_minimal()

```
```{r}
# Print counts by number of children
print("Children")
by_children <- table(insurance_data$children)
print(by_children)
cat(rep('-', 80), '\n')
```
## Analysis on categorical features.
categorical feature consist of sex, smoker, and region.
### analysis on sex
```{r} 

# Create a count plot for sex distribution
g <- ggplot(insurance_data, aes(x = sex)) +
  geom_bar(stat = "count", fill = "skyblue", alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count.., vjust = -0.2), alpha = 0.7) +
  ggtitle("Sex Distribution", subtitle = "Frequency/Count of Gender Type") +
  labs(x = "Sex", y = "Count")

# Display the plot
print(g)

```
```{r}
# Print counts by sex
print("Gender")
by_sex <- table(insurance_data$sex)
print(by_sex)
cat(rep('-', 80), '\n')
```
### analysis on smoker
```{r}
# count plot
# get freq/count
smoker_table <- table(data$smoker)
fr <- as.data.frame.table(smoker_table)
# plot freq
g <- ggplot(fr, aes(x="", y=Freq, fill = factor(Var1))) + 
  geom_bar(width=1, stat="identity") + 
  geom_text(aes(label = paste(round(Freq / sum(Freq) * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +
  ggtitle("smoker Distribution") + 
  labs(x="smoker", y="count") + 
  coord_polar(theta = "y", start=1)  + 
  theme(legend.position="bottom") + 
  guides(fill=guide_legend(title="")) + 
  scale_fill_manual(values = c("green", "red"))
 
p <- plot_grid(g)
title <- ggdraw() + draw_label("Analysis on smoker", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```
```{r}
# Print counts by smoker status
print("Smoker")
by_smoker <- table(insurance_data$smoker)
print(by_smoker)
cat(rep('-', 80), '\n')
```
There are less smoker. 

### analysis on region

```{r}
# count plot
# get freq/count
region_table <- table(data$region)
fr <- as.data.frame.table(region_table)
# plot freq
g <- ggplot(fr, aes(x="", y=Freq, fill = factor(Var1))) + 
  geom_bar(width=1, stat="identity") + 
  geom_text(aes(label = paste(round(Freq / sum(Freq) * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +
  ggtitle("region Distribution" ) + 
  labs(x="region", y="count") +
  coord_polar(theta = "y", start=1)  + 
  theme(legend.position="bottom") + 
  guides(fill=guide_legend(title="")) + 
  scale_fill_manual(values = c("#58508d", "#bc5090", "#ff6361", "#ffa600"))

p <- plot_grid(g)
title <- ggdraw() + draw_label("Analysis on region", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))

```
```{r}
# Print counts by region
print("Region")
by_region <- table(insurance_data$region)
print(by_region)
cat(rep('-', 80), '\n')
```
#Processing the dataset
## Encoding.

```{r}
encode <- function(x, order = unique(x)){
  x <- as.numeric(factor(x, levels = order, exclude = NULL))
  x
}
encoded_df <- insurance_data
encoded_df[["sex"]] <- encode(insurance_data[["sex"]])
encoded_df[["smoker"]] <- encode(insurance_data[["smoker"]])
encoded_df[["region"]] <- encode(insurance_data[["region"]])
head(encoded_df)
```

```{r}
str(encoded_df)
```
## correlation matrix


```{r}
#corr <- cor(encoded_df)
#g <- corrplot(corr, type="lower", tl.col = "black") 


# Select only numeric columns from the dataframe
numeric_columns <- sapply(encoded_df, is.numeric)
numeric_data <- encoded_df[, numeric_columns]

# Calculate correlation matrix
corr <- cor(numeric_data)

# Plot correlation matrix
library(corrplot)
g <- corrplot(corr, type = "lower", tl.col = "black")

```


```{r}
gg <- heatmap(x=corr, col = magma(10), symm=T)
```
# MODEL BUILDING

Spilt dataset to train and test with ratio 7:3.




```{r}
# Next, set a seed for reproducibility
set.seed(123)

# Split the data into a training set and a test set with a 70-30 split
split_index <- createDataPartition(encoded_df$charges, p = 0.7, list = FALSE)
train_data <- encoded_df[split_index, ]
test_data <- encoded_df[-split_index, ]

# Check the dimensions of the train set and test set
dim(train_data)
dim(test_data)


```



### Build Linear Regression Model
```{r}
lr <- lm(charges ~ age + sex + bmi + children + smoker + region, data=train_data)
print(lr)

# Predict on test data
test_data$pred <- predict(lr, newdata = test_data)

# Calculate R-squared
R_squared <- summary(lr)$r.squared
print(paste("R-squared:", R_squared))

```
# Print the accutracy for the Linear Regression Model
```{r}
R_squared
```


###  create a scatter plot of the actual and predicted values.
```{r}
gg <- ggplot(test_data, aes(x=pred, y=charges)) + 
  geom_point(color = "lightsalmon", alpha=0.8) + 
  geom_abline(color = "darkslategray", size=1) + 
  ggtitle("prediction vs real charges (Linear Regression")
gg
```
It's important to note that this graph can help you evaluate how well the model is performing by checking how closely the points are aligned to the line of perfect prediction. If the points are closely aligned to the line, the model is performing well. However, if the points are scattered away from the line, the model is not performing well and needs to be improved.
### model demonstration with dummy data

```{r}
dum <- data.frame(age = 24,
                  sex = 2, 
                  bmi = 30.1,
                  children = 0,
                  smoker = 1, 
                  region = 3)
print(paste0("Charges Prediction for dum: ", round(predict(lr, dum), 2)))
```

### Decision Tree model
```{r}
# build the decision tree 
decision_tr <- rpart(smoker ~ age + sex + bmi + region, data = train_data, method = "class")

# make predictions on the test data
predictions <- predict(decision_tr, test_data, type = "class")
print("-------------------------------------------------------------------------------------------")
# convert predictions and test data to factor and make the levels as same
predictions <- as.factor(predictions)
test_data$smoker <- as.factor(test_data$smoker)
levels(predictions) <- levels(test_data$smoker)

print("-------------------------------------------------------------------------------------------")
# evaluation of the the model 
# Confusion matrix to evalute Performance
confusion_matrix <- table(predicted = predictions, actual = test_data$smoker)
confusion_matrix
print("-------------------------------------------------------------------------------------------")
summary(decision_tr) 

```
#Accuracy of the Decision tree model
```{r}
accuracy_of_decision_tree <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("The accuracy of the Decision Tree model is : ", accuracy_of_decision_tree))
```
###build the Random forest model
```{r}
random_forest_model <- randomForest(as.factor(smoker) ~ ., data = train_data, ntree = 500)

# make predictions on the test data
random_forest_predictions <- predict(random_forest_model, newdata = test_data, type = "class")

print("-------------------------------------------------------------------------------------------")
# Evalute Performance
confusion_matrix <- table(predicted = random_forest_predictions, actual = test_data$smoker)
confusion_matrix
print("-------------------------------------------------------------------------------------------")

# Summary of the model
summary(random_forest_model)

```
# Calculate the accuracy of the model
```{r}
accuracy1 <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy of the Random Forest Model is:", accuracy1))
```
### Compare the accuracy of three model
```{r}
cat("Accuracy of Linear Model:", (R_squared)*100, "\n")
cat("Accuracy of Decision Tree Model:", (accuracy2)*100, "\n")
cat("Accuracy of Random Forest   Model:", (accuracy1)*100, "\n")
```



