---
title: "final_project"
author: "Diallo Mamadou Korka"
date: "2023-12-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the libraries required for this project
```{r }
library(ggplot2)
library(tidyverse)
library(readr)
library(corrplot)
library(cowplot)
library(viridis)
#library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```

# Load the data into Rstudio
```{r}
data <- read_csv("insurance.csv")
head(data)
```
# Check features in the dataset

```{r }
str(data)
```
# EXPLORATORY DATA ANALYSIS (EDA)
```{r}
summary(data)
```

## Check missing value.
```{r}
colSums(is.na(data))

```
## Analysis on categorical features.
categorical feature consist of sex, smoker, and region.



### analysis on sex
```{r}
# count plot
# get freq/count
sex_table <- table(data$sex)
fr <- as.data.frame.table(sex_table)
# plot freq
g <- ggplot(fr, aes(x=Var1, y=Freq)) + 
  geom_bar(stat="identity", fill="tomato2") + 
  geom_text(aes(label=Freq, vjust=-0.2), alpha=0.7) + 
  ggtitle("sex Distribution", subtitle = "frequency/count of gender type") + 
  labs(x="sex", y="count")
# scatter plot
gg <- ggplot(data, aes(x=sex, y=charges)) + 
  geom_jitter(aes(color=sex), alpha=0.7) + 
  ggtitle("sex vs charges")
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on sex", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```
Female vs male count/distribution balanced with slightly more male. There is no pattern on charges.
As we saw there is no missing data in our dataset.
### analysis on smoker
```{r}
# count plot
# get freq/count
smoker_table <- table(data$smoker)
fr <- as.data.frame.table(smoker_table)
# plot freq
g <- ggplot(fr, aes(x="", y=Freq, fill = factor(Var1))) + 
  geom_bar(width=1, stat="identity") + 
  geom_text(aes(label = paste(round(Freq / sum(Freq) * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +
  ggtitle("smoker Distribution", subtitle = "frequency/count of smoking customer") + 
  labs(x="smoker", y="count") + 
  coord_polar(theta = "y", start=1)  + 
  theme(legend.position="bottom") + 
  guides(fill=guide_legend(title="")) + 
  scale_fill_manual(values = c("#58508d", "#bc5090", "#ff6361", "#ffa600"))
# scatter plot
gg <- ggplot(data, aes(x=smoker, y=charges)) + 
  geom_jitter(color="lightcoral", alpha=0.7) + 
  ggtitle("smoker vs charges") + 
  theme(legend.position="None")
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on smoker", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```
There are less smoker. Customer who smoke show higher charges than those who don't.

### analysis on region

```{r}
# count plot
# get freq/count
region_table <- table(data$region)
fr <- as.data.frame.table(region_table)
# plot freq
g <- ggplot(fr, aes(x="", y=Freq, fill = factor(Var1))) + 
  geom_bar(width=1, stat="identity") + 
  geom_text(aes(label = paste(round(Freq / sum(Freq) * 100, 1), "%")),
            position = position_stack(vjust = 0.5)) +
  ggtitle("region Distribution", subtitle = "region of customer (US)") + 
  labs(x="region", y="count") +
  coord_polar(theta = "y", start=1)  + 
  theme(legend.position="bottom") + 
  guides(fill=guide_legend(title="")) + 
  scale_fill_manual(values = c("#58508d", "#bc5090", "#ff6361", "#ffa600"))
# scatter plot
gg <- ggplot(data, aes(x=region, y=charges)) + 
  geom_jitter(color="DarkOliveGreen", alpha=0.7) + 
  ggtitle("region vs charges") +
  theme(axis.text.x = element_text(angle=90), legend.position = "None") 
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on region", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))

```

There are 4 region, with 'southeast' has slightly more count than the other 3 that was pretty balanced to each other. There are no pattern on charges.
## Distribution analysis on numerical features.

Numerical features consist of age, bmi, children, and charges.

### distribution on age

```{r}
g <- ggplot(data, aes(age)) + 
  scale_fill_brewer(palette = "Spectral") + 
  geom_histogram(fill="thistle", binwidth = 1, col="black", size=.8) +
  ggtitle("age Distribution", subtitle = "distribution of age vs count")
gg <- ggplot(data, aes(x=age, y=charges)) + 
  geom_point(colour="RosyBrown", alpha=0.3) + 
  # geom_jitter(aes(color="re"), alpha=0.3) + 
  ggtitle("age vs charges") + 
  theme(legend.position="None")
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on age", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```

Ages have balanced distribution about 25-ish on age 20-64 with were more count on age 18 and 19. Age seems correspond linearly to charges. however, we can observe 3 cluster of charges, 0-20000, 20000ish, and 40000ish. 
### distribution on bmi
```{r}
g <- ggplot(data, aes(bmi)) + 
  geom_density(fill="#00798c", color="black", alpha=0.7) + 
  ggtitle("bmi Distribution", subtitle = "distribution of bmi vs count")
gg <- ggplot(data, aes(x=bmi, y=charges)) + 
  # geom_point(aes(col=bmi))
  geom_jitter(colour="#2e4057", alpha=0.3) + 
  ggtitle("bmi vs charges") + 
  theme(legend.position="None")
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on bmi", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```

Bmi distributed normally with average of 30.66. There are slightly increase of charges as the bmi increase.


### distribution on children count.

```{r}
g <- ggplot(data, aes(children)) + 
  scale_fill_brewer(palette = "Spectral") + 
  geom_histogram(fill="YellowGreen", binwidth = 1, col="black", size=.7) +
  ggtitle("children Distribution", subtitle = "distribution of number of children(dependents) covered")
gg <- ggplot(data, aes(x=children, y=charges)) + 
  # geom_point(aes(col=bmi))
  geom_jitter(colour = "lightseagreen", alpha=0.3) + 
  ggtitle("children vs charges") + 
  theme(legend.position="None")
p <- plot_grid(g, gg)
title <- ggdraw() + draw_label("Analysis on children", fontface = 'bold')
plot_grid(title, p, ncol = 1, rel_heights = c(0.1,1))
```

Most customer has no children, the more children count the less the counts. There are no pattern shown on charges.
## Encoding.

```{r}
encode <- function(x, order = unique(x)){
  x <- as.numeric(factor(x, levels = order, exclude = NULL))
  x
}
encoded_df <- data
encoded_df[["sex"]] <- encode(data[["sex"]])
encoded_df[["smoker"]] <- encode(data[["smoker"]])
encoded_df[["region"]] <- encode(data[["region"]])
head(encoded_df)
```

```{r}
str(encoded_df)
```
## correlation matrix


```{r}
corr <- cor(encoded_df)
g <- corrplot(corr, type="lower", tl.col = "black") 
```


```{r}
gg <- heatmap(x=corr, col = magma(10), symm=T)
```


As we expected, smoker has higher correlation to charges then other, there are also slight correlation to age and bmi.
# MODEL BUILDING

Spilt dataset to train and test with ratio 7:3.




```{r}
# Next, set a seed for reproducibility
set.seed(123)

# Split the data into a training set and a test set with a 70-30 split
split_index <- createDataPartition(encoded_df$charges, p = 0.7, list = FALSE)
train_data <- encoded_df[split_index, ]
test_data <- encoded_df[-split_index, ]

# Check the dimensions of the training and test sets
dim(train_data)
dim(test_data)


```



### Linear Regression Model
```{r}
lr <- lm(charges ~ age + sex + bmi + children + smoker + region, data=train_data)
print(lr)
# Predict on test data

test_data$pred <- predict(lr, newdata = test_data)

# Calculate R-squared
R_squared <- summary(lr)$r.squared
print(paste("R-squared:", R_squared))

```
# Print the accutracy for the Linear Regression Model
```{r}
R_squared
```


###  create a scatter plot of the actual and predicted values.
```{r}
gg <- ggplot(test_data, aes(x=pred, y=charges)) + 
  geom_point(color = "lightsalmon", alpha=0.8) + 
  geom_abline(color = "darkslategray", size=1) + 
  ggtitle("prediction vs real charges (Linear Regression")
gg
```
It's important to note that this graph can help you evaluate how well the model is performing by checking how closely the points are aligned to the line of perfect prediction. If the points are closely aligned to the line, the model is performing well. However, if the points are scattered away from the line, the model is not performing well and needs to be improved.
### model demonstration with dummy data

```{r}
dum <- data.frame(age = 24,
                  sex = 2, 
                  bmi = 30.1,
                  children = 0,
                  smoker = 1, 
                  region = 3)
print(paste0("Charges Prediction for dum: ", round(predict(lr, dum), 2)))
```

### Decision Tree
```{r}
# build the decision tree model
dr <- rpart(smoker ~ age + sex + bmi + region, data = train_data, 
               method = "class")

# make predictions on the test data
predictions <- predict(dr, test_data, type = "class")

# convert predictions and test data to factor and ensure they have the same levels
predictions <- as.factor(predictions)
test_data$smoker <- as.factor(test_data$smoker)
levels(predictions) <- levels(test_data$smoker)


# evaluate the model performance
# Evalute Performance
confusion_matrix <- table(predicted = predictions, actual = test_data$smoker)
confusion_matrix
summary(dr)
accuracy2 <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy of the Decision Tree model is:", accuracy2))

```
# Print accuracy for Decision tree model
```{r}
accuracy2 <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy of the Decision Tree model is:", accuracy2))
```








### Rnadom Forest
```{r}
# build the Random forest model
random_forest_model <- randomForest(as.factor(smoker) ~ ., data = train_data, ntree = 500)

# make predictions on the test data
random_forest_predictions <- predict(random_forest_model, newdata = test_data, type = "class")


# Evalute Performance
confusion_matrix <- table(predicted = random_forest_predictions, actual = test_data$smoker)
confusion_matrix


# Summary of the model
summary(random_forest_model)



```
# Calculate the accuracy of the model
```{r}
accuracy1 <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy of the Random Forest Model is:", accuracy1))
```
### Compare the accuracy of three model
```{r}
cat("Accuracy of Linear Model:", (R_squared)*100, "\n")
cat("Accuracy of Decision Tree Model:", (accuracy2)*100, "\n")
cat("Accuracy of Random Forest   Model:", (accuracy1)*100, "\n")
```



